Type-Safe Modular Parsing

### 2017-01-03 paper/sections/CaseStudy.tex
Revise section 6

To demonstrate the utility of our modular parsing library, we
implemented parsers of the first 18 calculi from book \textit{Types
  and Programming Languages} (TAPL). TAPL introduces several
calculi from simple to complex, by gradually adding new features to
syntax. \bruno{mention that you compare against an existing
  non-modular implementation.}

\subsection{Overview}\label{subsec:cs-overview}

The calculi in the TAPL book are suitable for our case study for mainly two reasons.
Firstly, the calculi are arguably capture many of the language features 
required in realistic programming languages. Secondly, the evolution of
calculi in the book reveals the advantages of modular representation
of abstract syntax and modular parsing, which is the key functionality
of our library. By decomposing those calculi into smaller components
and reusing the common ones, we can obtain considerably code reuse in
parsing with little performance penalty, as shown later.

\paragraph{Extracting Language Components}
Using the modular parsing techniques discussed before, we extract
reusable \textit{components} from syntax of all the calculi. Each
component, which may contain several syntactical structures, is a
reusable fragment representing a certain feature in the syntax. For
example, \lstinline{VarApp} component represents the variable case and
application case of expressions. In our case study, each component is
represented by a Scala object which has three parts. \lstinline{Alg}
is an object algebra interface for the abstract syntax.
\lstinline{Print} is an object algebra implements that interface
\lstinline{Alg}, for the pretty printing operation. It is not
necessary for parsing, but we include it here for demonstration.
\lstinline{Parser} is the modular parser for this piece of syntax.

\begin{lstlisting}
object VarApp {
  trait Alg[E] {
    def TmVar(x: String): E
    def TmApp(e1: E, e2: E): E
  }
  trait Print extends Alg[String] {
    def TmVar(x: String) = x
    def TmApp(e1: String, e2: String) = "[" + e1 + " " + e2 + "]"
  }
  trait Parser[E, F <: {val pE : PackratParser[E]}] {
    lexical.delimiters += ("(", ")")
    val pVarAppE: Alg[E] => (=> F) => PackratParser[E] = alg => l => {
      val e = l.pE
      ident ^^ alg.TmVar |||
      e ~ e ^^ { case e1 ~ e2 => alg.TmApp(e1, e2) } |||
      "(" ~> e <~ ")"
    }
  }
}
\end{lstlisting}

We use some naming conventions in our code, \lstinline{E} represents
expressions, \lstinline{T} represents types and \lstinline{K}
represents kinds. In this component \lstinline{VarApp} we only have
expressions. The \lstinline{Alg} and \lstinline{Print} are standard
object algebra operations, thus we will focus on the modular parser
\lstinline{Parser}. Each parser carries lexical information and
parsing functions. The lexical information includes reserved words,
delimiters, etc. Parsing functions are similar with we described
before, except that we use record types under some subtyping
constraints here for the explicit self-reference. The record type
contains a parser for expressions \lstinline{pE}, a parser for types
\lstinline{pT} and a parser for kinds \lstinline{pK}, if the
corresponding one is needed when parsing the syntax of this component.
Here the type of self-reference, namely \lstinline{F}, is only
required to have a parser of expressions \lstinline{pE}. Then we can
build the parsing function \lstinline{pVarAppE}, the suffix
\lstinline{E} also denotes that it will parse an expression.

\paragraph{Composing Language Components}
Each calculus could be composed directly from components and other
calculi, if it has those common structures in the syntax. For example,
the calculus \lstinline{Untyped}, representing the famous untyped
lambda calculus, can be constructed from components \lstinline{VarApp}
and untyped lambda abstraction \lstinline{UntypedAbs}.

\begin{lstlisting}
object UntypedAbs {
  trait Alg[E] {
    def TmAbs(x: String, e: E): E
  }
  trait Print extends Alg[String] {
    def TmAbs(x: String, e: String) = "\\" + x + "." + e
  }
  trait Parser[E, F <: {val pE : PackratParser[E]}] {
    val pUntypedAbsE: Alg[E] => (=> F) => PackratParser[E] = {...}
  }
}

object Untyped {
  trait Alg[E] extends UntypedAbs.Alg[E] with VarApp.Alg[E]
  trait Print extends Alg[String] with UntypedAbs.Print with VarApp.Print
  trait Parser[E, L <: {val pE : PackratParser[E]}] extends UntypedAbs.Parser[E, L] with VarApp.Parser[E, L] {
    val pUntypedE = pUntypedAbsE | pVarAppE
  }
}
\end{lstlisting}

When composing the calculus from components, all the fields can be
easily combined by Scala’s \lstinline{extends...with} keywords as
shown in the code. For \lstinline{Parser}, the subtyping constraint
for the explicit self-reference should satisfy all the requirement of
the parsers it extends from. \huang{Talk about | here?}

Finally, we have demo code for every calculus. The parsing result is a
string because we provide the pretty printing algebra to the parser,
and it can be changed as long as the concrete algebra is available. In
this case study we only have pretty printing as the operation, since
it is enough for demonstration.

\begin{lstlisting}
object TestUntyped {
    to be polished
}
\end{lstlisting}

In the next section we will show more examples and talk further about
how we extend and compose these components and calculi.

\subsection{Extensibility}\label{subsec:cs-extensibility}

@@

As mentioned before, we decompose all the calculi into reusable
components for modularity. Figure \ref{fig:dependency} shows the
dependency of all the components and calculi in our case study. Gray
boxes are calculi and white boxes are components. An arrow starting
from box A to box B denotes that B includes and thus reuses A.

As shown in the graph, some components such as \lstinline{VarApp} are
created from scratch, while others such as \lstinline{Typed} are
extended from existing components. The \lstinline{Typed} component
extends from \lstinline{VarApp} and has two more syntactical
structures which are typed lambda abstraction and function type (arrow
type). It is the core of many calculi in this case study, and we use
it here as an example for demonstrating the extensibility of
components.

\begin{lstlisting}
object Typed {
  trait Alg[E, T] extends VarApp.Alg[E] {
    def TmAbs(x: String, t: T, e: E): E
    def TyArr(t1: T, t2: T): T
  }
  trait Print extends Alg[String, String] with VarApp.Print {
    def TmAbs(x: String, t: String, e: String) = "\\(" + x + ":" + t + ")." + e
    def TyArr(t1: String, t2: String) = t1 + "->" + t2
  }
  trait Parser[E, T, F <: {val pE : PackratParser[E]; val pT : PackratParser[T]}] extends VarApp.Parser[E, F] {
    ...
    private val pAbsE: Alg[E, T] => (=> F) => PackratParser[E] = {...}
    val pTypedE: Alg[E, T] => (=> F) => PackratParser[E] = pVarAppE | pAbsE
    val pTypedT: Alg[E, T] => (=> F) => PackratParser[T] = alg => l => {
      val t = l.pT
      t ~ ("->" ~> t) ^^ { case t1 ~ t2 => alg.TyArr(t1, t2) } ||| "(" ~> t <~ ")"
    }
  }
}
\end{lstlisting}

Except building a component by extending an existing one, here we have
another extension which is adding a new sort of syntax. It has been
discussed in section \ref{subsec:openrecursion}. In this example,
originally \lstinline{VarApp} has only one sort of syntax which is
expressions, and then \lstinline{Typed} introduces types to the
syntax. We add an extra type parameter \lstinline{T} for it, so that
it is able to be distinguished. The \lstinline{Parser} also requires
the self-reference to have a field \lstinline{pT} for parsing types,
which is represented by the subtyping constraint. With this support,
the \lstinline{pTypedT} function parses arrow types using
\lstinline{pT}. Another point worth mentioning is that we use a
private local function \lstinline{pAbsE} here for parsing the typed
lambda abstraction case, and we can directly combine it with the
inherited \lstinline{pVarApp} to obtain our parsing function of
expressions.

Since calculi and components have the similar signature, each calculus
can also be extended and reused directly. Our last example shows how
to build calculus \lstinline{FullRef} from extending
\lstinline{FullSimple}, another calculus.

\begin{lstlisting}
object FullRef {
  trait Alg[E, T] extends FullSimple.Alg[E, T] with TopBot.Alg[T] with Ref.Alg[E, T]
  trait Print extends Alg[String, String] with FullSimple.Print with TopBot.Print with Ref.Print
  trait Parser[E, T, L <: {val pE : PackratParser[E]; val pT : PackratParser[T]}] extends FullSimple.Parser[E, T, L] with TopBot.Parser[T, L] with Ref.Parser[E, T, L] {
    val pFullRefE = pFullSimpleE | pRefE
    val pFullRefT = pFullSimpleT | pRefT | pTopBotT
  }
}
\end{lstlisting}

Using our modular parsing library, dividing the syntax into small
pieces and bind them with corresponding parsers to form reusable
language component is convenient. From the dependency graph we know
that the common components such as \lstinline{VarApp} are reused in
lots of calculi. Such reuse could shorten the code considerably. We
will show this advantage and exam the possible performance penalty in
a quantitative way in the next section.

@@

We compared our implementation of the parsers with an implementation
by  Ilya Klyuchnikov, available online. Ilya's
implementation is suitable for comparison, because it is also
written in Scala using the same parser combinator library.
Furthermore, it includes parsers of all the 18 calculi we have, but
written in a non-modular way. Thus it is not able to reuse existing
code of parsing common syntactical features.

Because Ilya's implementation has extra code such as semantics of
those calculi, we removed all irrelevant code and only keep abstract
syntax definition, parser and pretty printer for each calculus, to
ensure a fair comparison.

The comparison is made from two aspects. First, we want to show that
using our modular parsing library, the code can be reused in a
significant way. For this propose, we measured source lines of code
(SLOC) of the two implementations. All blank lines and comment lines
are excluded, and we formatted the code of both implementations so
that the length of each line does not exceed 120 characters. Second,
we are interested to assess the performance penalty caused by modularity. Thus
we compared the execution time, as a major measurement of performance,
between the two implementations. We prepared test files containing
valid expressions in each calculus,\bruno{say something about what
  those expressions are, size, and how they were generated.} then we used the corresponding
parser to parse the file and the pretty printer to print the result.
The execution time excluding reading input file was calculated.

@@

Table \ref{tab:comparison} shows data of the comparison. The overall
result of SLOC comparison is that 66.3\% of code is reduced using our
library. That is because the code of some common syntax structures
such as variables, lambda abstractions, etc., are reused several times in
the whole project. We can see that in the first two calculi
\lstinline{Arith} and \lstinline{Untyped} we do worse than Ilya's
implementation, because in such two cases we do not reuse any existing
components. But in each of the following 16 calculi, we can compose
our syntax definition, parser and pretty printer from extending
reusable components, resulting considerably code reduction.

\paragraph{Performance}
For the execution time comparison, we built a generator to randomly
generate valid expressions according to the syntax of each calculus.
Each test file corresponds with one calculus and it consists of 500
expressions randomly generated. The size of test files varies from
20KB to 100KB. After testing both implementation, we found that our
version is about \huang{60\%} slower. \bruno{A table showing a summary
of results is missing here. We discussed this before.}
 Part of the reason is that our
library requires more intermediate function calls, and the object
algebra and self-references need to be passed as arguments every
time. However, another important reason is that, because of the
modular setting, we use a different alternative combinator which will
try both branches and match the longest one. \huang{ref to somewhere
  before?} In contrast Ilya uses the ordinary alternative combinator
which tries the second branch only if the first one fails. Although
this difference slows our version, we want to mention that because of
the memoization of Packrat parsers, we are just constant times
slower, the big-O complexity is still the same.

>>>>

To demonstrate the utility of our modular parsing approach, we
implemented parsers of the first 18 calculi from book \textit{Types
  and Programming Languages} (TAPL) []. TAPL introduces several
calculi from simple to complex, by gradually adding new features to
syntax. We compared our implementation with a non-modular implementation we found online [], which is also written in Scala and uses the same Packrat parsing library. The comparison is made from two aspects, source lines of code (SLOC) and execution time. \huang{done}\bruno{mention that you compare against an existing
  non-modular implementation.}

\subsection{Implementation}

The calculi in the TAPL book are suitable for our case study for mainly two reasons.
Firstly, the calculi are arguably capture many of the language features
required in realistic programming languages. Secondly, the evolution of
calculi in the book reveals the advantages of modular representation
of abstract syntax and modular parsing, which is the key functionality
of our approach. By extracting common components from those calculi
and reusing them, we obtain considerably code reuse with reasonable performance penalty, as shown later.

\paragraph{Extracting Language Components}
Using the modular parsing techniques discussed before, we extract
reusable components from syntax of all the calculi. Each
component, which may contain several syntactical structures,
represents a certain feature of the language. For
example, the \lstinline{VarApp} component below represents variables and
function applications.

\huang{todo: move the code, rename PackratParser}
\begin{lstlisting}
object VarApp {
  trait Alg[E] {
    def TmVar(x: String): E
    def TmApp(e1: E, e2: E): E
  }
  trait Print extends Alg[String] {
    def TmVar(x: String): String = x
    def TmApp(e1: String, e2: String): String = "[" + e1 + " " + e2 + "]"
  }
  trait Parser[E] extends EParser[E] {
    lexical.delimiters += ("(", ")")
    val alg: Alg[E]
    val pVarAppE: PackratParser[E] = {
      lcid ^^ alg.TmVar |||
        pE ~ pE ^^ { case e1 ~ e2 => alg.TmApp(e1, e2) } |||
        "(" ~> pE <~ ")"
    }
  }
}
\end{lstlisting}

In our case study, each component is represented by a Scala object which has three parts.
\lstinline{Alg} is an object algebra interface for the abstract syntax.
\lstinline{Print} is an object algebra implements that interface
\lstinline{Alg}, for the pretty printing operation. They are standard object algebra operations.
\lstinline{Parser} is the modular parser for this piece of syntax.

We have some naming conventions in our code, \lstinline{E} represents
expressions, \lstinline{T} represents types and \lstinline{K}
represents kinds. They are the three sorts of syntax in our case study.
In this particular component \lstinline{VarApp} we only have expressions.

Similar as before, each parser carries lexical information, an object algebra field, and
parsing functions. We have some helper traits for convenience, such as \inlinecode{EParser}
containing \inlinecode{pE} for parsing expressions.

\begin{lstlisting}
trait EParser[E] {
  val pE: PackratParser[E]
}
\end{lstlisting}

\paragraph{Composing Language Components}
Each calculus could be composed directly from components and other
calculi as needed. For example, the calculus \lstinline{Untyped} in our case study,
representing the famous untyped lambda calculus, is constructed from components \lstinline{VarApp}
and untyped lambda abstraction \lstinline{UntypedAbs}.

\begin{lstlisting}
object UntypedAbs {
  trait Alg[E] {
    def TmAbs(x: String, e: E): E
  }
  trait Print extends Alg[String] {
    def TmAbs(x: String, e: String): String = "\\" + x + "." + e
  }
  trait Parser[E] extends EParser[E] {
    lexical.delimiters += ("\\", ".", "(", ")")
    val alg: Alg[E]
    val pUntypedAbsE: PackratParser[E] =
      ("\\" ~> lcid) ~ ("." ~> pE) ^^ { case x ~ e0 => alg.TmAbs(x, e0) } |||
      "(" ~> pE <~ ")"
  }
}

object Untyped {
  trait Alg[E] extends UntypedAbs.Alg[E] with VarApp.Alg[E]
  trait Print extends Alg[String] with UntypedAbs.Print with VarApp.Print
  trait Parser[E] extends UntypedAbs.Parser[E] with VarApp.Parser[E] {
    override val alg: Alg[E]
    val pUntypedE: PackratParser[E] = pUntypedAbsE ||| pVarAppE
    override val pE: PackratParser[E] = pUntypedE
  }
}
\end{lstlisting}

When composing the calculus from components, all the traits can be
easily combined by Scala's \inlinecode{extends...with} keywords.
In \lstinline{Parser}, we need to override the object algebra interface and
parsing functions accordingly.

\paragraph{Demo code}
We have demo code for every calculus in order to test and evaluate our implementation.
We feed the pretty printing algebra to the parser and print out the result.
The parsing result can be changed as long as the concrete algebra is available.

\begin{lstlisting}
object TestUntyped {
  def work[E](inp: String)(a: Untyped.Alg[E]): E = {
    val lang = new Untyped.Parser[E] {
      override val alg: Untyped.Alg[E] = a
    }
    parse(lang.pE)(inp)
  }
  def parseAndPrint(inp: String): Unit = println(work(inp)(new Untyped.Print {}))
  def main(args: Array[String]): Unit = {
    parseAndPrint("(\\x.x) (\\y.y)")
  }
}
\end{lstlisting}

\paragraph{Dependency Overview}
Figure \ref{fig:dependency} shows the
dependency of all the components and calculi in our case study. Gray
boxes are calculi and white boxes are components. An arrow starting
from box A to box B denotes that B includes and thus reuses A.

@@

As shown in the graph, some components such as \lstinline{VarApp} are
created from scratch, while others such as \lstinline{Typed} are
extended from existing components. Since calculi and components have the similar signature, each calculus
can also be extended and reused directly. For example, calculus \lstinline{FullRef} extends from
calculus \lstinline{FullSimple}.

Using our modular parsing approach, it is convenient to extract language
features as components and reuse them. From the dependency graph, we know
that common components such as \lstinline{VarApp} are reused in
lots of calculi. Such reuse could shorten the code considerably. We
will show this advantage and exam the possible performance penalty in
a quantitative way in the next subsection.

@@

We compared our implementation of the parsers with an implementation
by Ilya Klyuchnikov, available online []. Ilya's
implementation is suitable for comparison, because it is also
written in Scala using the same parser combinator library.
Furthermore, it includes parsers of all the 18 calculi we have, but
written in a non-modular way. Thus it is not able to reuse existing
code when those calculi share common features.

The comparison is made from two aspects. First, we want to discover
the amount of code reuse using our modular parsing approach.
For this propose, we measured source lines of code (SLOC) of two implementations.
Second, we are interested to assess the performance penalty caused by modularity.
Thus we compared the execution time of parsing random expressions between two implementations.

\paragraph{Standard of Comparison}
In the SLOC comparison, all blank lines and comment lines are excluded,
and we formatted the code of both implementations to guarantee that
the length of each line does not exceed 120 characters. Furthermore,
because Ilya's implementation has extra code such as semantics,
we removed all irrelevant code and only keep abstract
syntax definition, parser and pretty printer for each calculus, to
ensure a fair comparison.

For the comparison of execution time, we built a generator to randomly
generate valid expressions of each calculus, according to the syntax. These expressions are
written to test files, one file per calculus. Each test file consists of 500
expressions randomly generated, and the size of test files varies from 20KB to 100KB.
Then we run the corresponding parser to parse the file and the pretty printer to print the result.
The average execution time of 5 runs excluding reading input file was calculated, in milliseconds.

\huang{done}\bruno{say something about what those expressions are, size, and how they were generated.}

@@

\paragraph{Comparison Results}
Table \ref{tab:comparison} shows results of the comparison.
The overall result is that 69.2\% of code is reduced using our
library, and our implementation is 42.7\% slower.

The good SLOC result is because of that the code of common language features
such as variables, lambda abstractions, etc., are reused lots of times in
the whole case study. We can see that in the first two calculi
\lstinline{Arith} and \lstinline{Untyped} we are not better than Ilya's
implementation, because in such two cases we do not reuse any existing
components. However in the following 16 calculi, we can reuse
language components, resulting considerably code reduction. In particular,
the calculi \inlinecode{EquiRec}, \inlinecode{Recon} and some others are only 22 lines
in our implementation, because we only compose existing codes in such cases.

To discover the reasons of slower execution time, we made some more experiments
on two possible factors which could affect the performance.
They are object algebra and the longest matching alternative combinator.
We use object algebra for ASTs and the longest matching alternative combinator \inlinecode{|||} for parsing,
while Ilya's implementation uses case class and the ordinary alternative combinator.

We implemented two more versions. One is a modified version of our implementation,
with object algebra replaced by case class for the ASTs.
The other is a modified version of Ilya's implementation, using the longest matching
alternative combinator instead of the ordinary one.

\huang{done}\bruno{A table showing a summary of results is missing here. We discussed this before.}

\begin{table}
    \centering
    \begin{tabular}{|l|r|r|r|r|r|r|r|}
      \hline
        \multirow{2}{*}{\bfseries Calculus Name} & \bfseries Ilya's & \multicolumn{2}{ c| }{\bfseries Ours} & \multicolumn{2}{ c| }{\bfseries Ilya's with \inlinecode{|||}} & \multicolumn{2}{ c| }{\bfseries Ours with class} \\ \cline{3-8}
        \multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{\bfseries Time} & \bfseries Time & \bfseries (+/-)\% & \bfseries Time & \bfseries (+/-)\% & \bfseries Time & \bfseries (+/-)\% \\
      \hline
        Arith & 741 & 913 & +23.2 & 793 & +7.0 & 932 & +25.8 \\
        Untyped & 770 & 1018 & +32.2 & 821 & +6.6 & 1007 & +30.8 \\
      \hline
        \multicolumn{1}{|c|}{\dots} & \multicolumn{7}{c|}{\dots} \\
      \hline
        FullRecon & 1094 & 1645 & +50.4 & 1161 & +6.1 & 1652 & +51.0 \\
        FullPoly & 1398 & 2086 & +49.2 & 1511 & +8.1 & 2019 & +44.4 \\
        FullOmega & 1451 & 2352 & +62.1 & 1582 & +9.0 & 2308 & +59.1 \\
      \hline
        Total & 21746 & 31024 & +42.7 & 23260 & +7.0 & 30795 & +41.6 \\
      \hline
    \end{tabular}
    \caption{Execution time of four versions}
    \label{tab:ext-comparison}
\end{table}

Table \ref{tab:ext-comparison} shows part of the comparison results of these four versions,
full table is available online []. It suggests that the difference of running time between
using object algebra and class is little, roughly 1\%.
The usage of longest matching combinator slows the performance by 7\%. The main reason of slower
execution time may be the overall structure of the modular parsing approach, because we indeed have
more intermediate function calls and method overriding. However, it is worth metioning that
because of the memoization technique of Packrat parsers, we are only constant times
slower, the algorithmic complexity is still the same.
